---
title: "NHIS"
author: "Colin Heneeman"
subtitle: "Correlates of Mental Health Using NHIS Data"
format: pdf
editor: source
editor_options: 
  chunk_output_type: console
---

#Prep

```{r}
#load packages
library(readr)
library(dplyr)
library(tidyverse)
library(haven)
library(psych)
library(car)
library(descr)
library(DescTools) 
library(ggpubr) 
library(rstatix)
library(DescTools)
library(nortest) 
library(emmeans)
library(ggplot2)
library(stargazer)
library(corrplot)
library(coefplot)
```

```{r, echo = FALSE}
# read original dataset - adult22 because it is the 2022 NHIS dataset

adult22 <- read_csv("adult22csv/adult22.csv")

glimpse(adult22)
```

##Create Smaller Dataset

```{r}

# create smaller dataset

adult22_cleaned <- select(adult22,
                          PHSTAT_A, LSATIS4_A, ANXFREQ_A, ANXLEVEL_A, DEPFREQ_A, DEPLEVEL_A, 
                          GAD71_A, GAD72_A, GAD73_A, GAD74_A, GAD75_A, GAD76_A, GAD77_A, PA18_02R_A, 
                          PHQ81_A, PHQ82_A, PHQ83_A, PHQ84_A, PHQ85_A, PHQ86_A, PHQ87_A, PHQ88_A, 
                          MODTPR_A, FRIESTP_A, VIGTPR_A, VIGFREQW_A, SLPHOURS_A, SLPREST_A, 
                          SLPFLL_A, SLPSTY_A, OVEGTP_A, SODATP_A, FRUITTP_A, GADCAT_A, PHQCAT_A, 
                          SEX_A, AGEP_A, HISPALLP_A, SALADTP_A, DRK12MWK_A, SMKNOW_A, CIGNOW_A, 
                          ECIGNOW_A, EMPWKHRS3_A)



# renaming to NHIS for simplicity 
NHIS <- adult22_cleaned



```

# Manipulate Variables

In this section, I will start to re-code and manipulate the variables of interest from the dataset to prepare for understanding correlations, model-building, and finally hypothesis testing.

## Dependent Variables

Below, each section is developed for one of the dependent variables of interest here.

### Life Satisfaction

```{r}
# life satisfaction - LSATIS4_A

# ITEM LSATIS4_A : 'In general, how satisfied are you with your life? Are you very satisfied, satisfied, dissatisfied, or very dissatisfied? 

# CODE: (Scaled 1 ~ very satisfied to 4 ~ very dissatisfied)

# checking initial values
table(NHIS$LSATIS4_A)

# re-scaling so that higher values = more satisfaction
NHIS <- NHIS %>%
  mutate(LSATIS4_A = as.numeric(LSATIS4_A),
         LSATIS4_A = case_when(
           LSATIS4_A == 1 ~ 4,
           LSATIS4_A == 2 ~ 3,
           LSATIS4_A == 3 ~ 2,
           LSATIS4_A == 4 ~ 1,
           LSATIS4_A == 7 ~ NA,
           LSATIS4_A == 8 ~ NA,
           LSATIS4_A == 9 ~ NA,
           TRUE ~ LSATIS4_A))

# checking to confirm value changes
table(NHIS$LSATIS4_A)


```

### Anxiety Measures

```{r}

# Measure 1 : Anxiety Frequency
 # ITEM: ANXFREQ_A - How often do you feel anxious?
 # CODE - 1 - Daily : 5 - Never

table(NHIS$ANXFREQ_A)

# Reversing scale such that higher frequency = higher score
NHIS <- NHIS %>%
  mutate(ANXFREQ_A = as.numeric(ANXFREQ_A),
         ANXFREQ_A = case_when(
           ANXFREQ_A == 1 ~ 5,
           ANXFREQ_A == 2 ~ 4,
           ANXFREQ_A == 3 ~ 3,
           ANXFREQ_A == 4 ~ 2,
           ANXFREQ_A == 5 ~ 1,
           ANXFREQ_A == 7 ~ NA,
           ANXFREQ_A == 8 ~ NA,
           ANXFREQ_A == 9 ~ NA,
           TRUE ~ ANXFREQ_A))

table(NHIS$ANXFREQ_A)

summary(NHIS$ANXFREQ_A)


# Measure 2 - GAD-7 (Generalized Anxiety Disorder)

# GAD-7 anxiety : a 7 item questionnaire typically used for diagnostic purposes of Generalized Anxiety Disorder 

# CODED : 1 - Not at all to 4 - Nearly every day

# ITEM: GAD71_A : Over the last two weeks, how often have you been bothered by the following problems - feeling nervous, anxious, or on edge 



table(NHIS$GAD71_A)

NHIS <- NHIS %>%
  mutate(GAD71A = as.numeric(GAD71_A),
         GAD71_A = case_when(
           GAD71_A== 1 ~ 1,
           GAD71_A == 2 ~ 2,
           GAD71_A== 3 ~ 3,
           GAD71_A == 4 ~ 4,
           GAD71_A == 7 ~ NA,
           GAD71_A == 8 ~ NA,
           GAD71_A == 9 ~ NA,
           TRUE ~ GAD71_A))

table(NHIS$GAD71_A)

# GAD72_A : Not being able to stop or control worrying
# 1 - Not at all, 2 - Several Days, 3 - More than half the days, 4 - Nearly Every day

table(NHIS$GAD72_A)

NHIS <- NHIS %>%
  mutate(GAD72_A = as.numeric(GAD72_A),
         GAD72_A = case_when(
           GAD72_A== 1 ~ 1,
           GAD72_A == 2 ~ 2,
           GAD72_A== 3 ~ 3,
           GAD72_A == 4 ~ 4,
           GAD72_A == 7 ~ NA,
           GAD72_A == 8 ~ NA,
           GAD72_A == 9 ~ NA,
           TRUE ~ GAD72_A))

table(NHIS$GAD72_A)

# GAD73_A
# Over the last two weeks, how often have you been bothered by ... worrying too much about different things - 

# How often worrying too much, past 2 weeks

# Code : 1 - Not at all , 2 - Several days , 3 - More than half the days , 4 - Nearly every day

table(NHIS$GAD73_A)

NHIS <- NHIS %>%
  mutate(GAD73_A = as.numeric(GAD73_A),
         GAD73_A = case_when(
           GAD73_A== 1 ~ 1,
           GAD73_A == 2 ~ 2,
           GAD73_A== 3 ~ 3,
           GAD73_A == 4 ~ 4,
           GAD73_A == 7 ~ NA,
           GAD73_A == 8 ~ NA,
           GAD73_A == 9 ~ NA,
           TRUE ~ GAD73_A))

table(NHIS$GAD73_A)

# GAD74_A
# Over the last two weeks, how often have you been bothered by .... 

# Trouble relaxing

# Code: 1 - Not at all, 2 - Several days, 3 - More than half the days, 4 - Nearly every day


table(NHIS$GAD74_A)

NHIS <- NHIS %>%
  mutate(GAD74_A = as.numeric(GAD74_A),
         GAD74_A = case_when(
           GAD74_A== 1 ~ 1,
           GAD74_A == 2 ~ 2,
           GAD74_A== 3 ~ 3,
           GAD74_A == 4 ~ 4,
           GAD74_A == 7 ~ NA,
           GAD74_A == 8 ~ NA,
           GAD74_A == 9 ~ NA,
           TRUE ~ GAD74_A))

table(NHIS$GAD74_A)

# GAD75_A
# Over the last two weeks, how often have you been bothered by

# Being so restless that it's hard to sit still

# Code: 1 - Not at all, 2 - Several days, 3 - More than half the days, 4 - Nearly every day

table(NHIS$GAD75_A)

NHIS <- NHIS %>%
  mutate(GAD75_A = as.numeric(GAD75_A),
         GAD75_A = case_when(
           GAD75_A== 1 ~ 1,
           GAD75_A == 2 ~ 2,
           GAD75_A== 3 ~ 3,
           GAD75_A == 4 ~ 4,
           GAD75_A == 7 ~ NA,
           GAD75_A == 8 ~ NA,
           GAD75_A == 9 ~ NA,
           TRUE ~ GAD75_A))

table(NHIS$GAD75_A)

# GAD76_A
# Over the last two weeks, how often have you been bothered by...

# Becoming easily annoyed or irritable

# Code: 1 - Not at all, 2 - Several days, 3 - More than half the days, 4 - Nearly every day

table(NHIS$GAD76_A)

NHIS <- NHIS %>%
  mutate(GAD76_A = as.numeric(GAD76_A),
         GAD76_A = case_when(
           GAD76_A== 1 ~ 1,
           GAD76_A == 2 ~ 2,
           GAD76_A== 3 ~ 3,
           GAD76_A == 4 ~ 4,
           GAD76_A == 7 ~ NA,
           GAD76_A == 8 ~ NA,
           GAD76_A == 9 ~ NA,
           TRUE ~ GAD76_A))

table(NHIS$GAD76_A)

 # GAD77_A
table(NHIS$GAD77_A)

# Over the last two weeks, how often have you been bothered by..

# Feeling afraid as if something awful might happen?

# Code: 1 - Not at all, 2 - Several days, 3 - More than half the days, 4 - Nearly every day

NHIS <- NHIS %>%
  mutate(GAD77_A = as.numeric(GAD77_A),
         GAD77_A = case_when(
           GAD77_A== 1 ~ 1,
           GAD77_A == 2 ~ 2,
           GAD77_A== 3 ~ 3,
           GAD77_A == 4 ~ 4,
           GAD77_A == 7 ~ NA,
           GAD77_A == 8 ~ NA,
           GAD77_A == 9 ~ NA,
           TRUE ~ GAD77_A))

table(NHIS$GAD77_A)

# GADCAT_A
# Severity of anxiety symptoms 

# Code: 1 - None/Minimal, 2 - Mild, 3 - Moderate, 4 - Severe, 8 - Not Ascertained

table(NHIS$GADCAT_A)

NHIS <- NHIS %>%
  mutate(GADCAT_A = as.numeric(GADCAT_A),
         GADCAT_A = case_when(
           GADCAT_A== 1 ~ 1,
           GADCAT_A == 2 ~ 2,
           GADCAT_A== 3 ~ 3,
           GADCAT_A == 4 ~ 4,
           GADCAT_A == 7 ~ NA,
           GADCAT_A == 8 ~ NA,
           GADCAT_A == 9 ~ NA,
           TRUE ~ GADCAT_A))

table(NHIS$GADCAT_A)

# Calculating a GAD 'Score' for each individual to use as a measure for anxiety - essentially taking an individual's average on each of the 7 measures of GAD 

NHIS$GADSCORE <- rowMeans(NHIS[, c("GAD71_A", "GAD72_A", "GAD73_A", "GAD74_A", "GAD75_A", "GAD76_A", "GAD77_A")], na.rm = TRUE)

summary(NHIS$GADSCORE)

```

### Depression

```{r}

 # DEPFREQ_A : How often do you feel depressed? Would you say daily, weekly, monthly, a few times a year, or never?

table(NHIS$DEPFREQ_A)

# reversing scale
NHIS <- NHIS %>%
  mutate(DEPFREQ_A = as.numeric(DEPFREQ_A),
         DEPFREQ_A = case_when(
           DEPFREQ_A== 1 ~ 5,
           DEPFREQ_A == 2 ~ 4,
           DEPFREQ_A == 3 ~ 3,
           DEPFREQ_A == 4 ~ 2,
           DEPFREQ_A == 5 ~ 1,
           DEPFREQ_A == 7 ~ NA,
           DEPFREQ_A == 8 ~ NA,
           DEPFREQ_A == 9 ~ NA,
           TRUE ~ DEPFREQ_A))

table(NHIS$DEPFREQ_A)

summary(NHIS$DEPFREQ_A)

# Measure 2 - PHQ-8 Depression / Major Depressive Disorder

# A similar 8 item questionnaire used to assess depression.

 # CODE: 1 - Not at all, 2 - Several Days, 3 - More than half the days, 4 - Nearly every day

table(NHIS$PHQ81_A)

NHIS <- NHIS %>%
  mutate(PHQ81_A = as.numeric(PHQ81_A),
         PHQ81_A = case_when(
           PHQ81_A== 1 ~ 1,
           PHQ81_A == 2 ~ 2,
           PHQ81_A== 3 ~ 3,
           PHQ81_A == 4 ~ 4,
           PHQ81_A == 7 ~ NA,
           PHQ81_A == 8 ~ NA,
           PHQ81_A == 9 ~ NA,
           TRUE ~ PHQ81_A))

table(NHIS$PHQ81_A)

 # PHQ82_A

 #feeling down, depressed, or hopeless

 # 1 - Not at all, 2 - Several Days, 3 - More than half the days, 4 - Nearly every day

table(NHIS$PHQ82_A)

NHIS <- NHIS %>%
  mutate(PHQ82_A = as.numeric(PHQ82_A),
         PHQ82_A = case_when(
           PHQ82_A== 1 ~ 1,
           PHQ82_A == 2 ~ 2,
           PHQ82_A== 3 ~ 3,
           PHQ82_A == 4 ~ 4,
           PHQ82_A == 7 ~ NA,
           PHQ82_A == 8 ~ NA,
           PHQ82_A == 9 ~ NA,
           TRUE ~ PHQ82_A))

table(NHIS$PHQ82_A)

 # PHQ83_A

 # trouble falling or staying asleep, or sleeping too much

 # 1 - Not at all, 2 - Several Days, 3 - More than half the days, 4 - Nearly every day

table(NHIS$PHQ83_A)

NHIS <- NHIS %>%
  mutate(PHQ83_A = as.numeric(PHQ83_A),
         PHQ83_A = case_when(
           PHQ83_A== 1 ~ 1,
           PHQ83_A == 2 ~ 2,
           PHQ83_A== 3 ~ 3,
           PHQ83_A == 4 ~ 4,
           PHQ83_A == 7 ~ NA,
           PHQ83_A == 8 ~ NA,
           PHQ83_A == 9 ~ NA,
           TRUE ~ PHQ83_A))

table(NHIS$PHQ83_A)

 # PHQ84_A

 # feeling tired or having little energy
 # 1 - Not at all, 2 - Several Days, 3 - More than half the days, 4 - Nearly every day

table(NHIS$PHQ84_A)

NHIS <- NHIS %>%
  mutate(PHQ84_A = as.numeric(PHQ84_A),
         PHQ84_A = case_when(
           PHQ84_A== 1 ~ 1,
           PHQ84_A == 2 ~ 2,
           PHQ84_A== 3 ~ 3,
           PHQ84_A == 4 ~ 4,
           PHQ84_A == 7 ~ NA,
           PHQ84_A == 8 ~ NA,
           PHQ84_A == 9 ~ NA,
           TRUE ~ PHQ84_A))

table(NHIS$PHQ84_A)


 # PHQ85_A

 # poor appetite or overeating
 # 1 - Not at all, 2 - Several Days, 3 - More than half the days, 4 - Nearly every day

table(NHIS$PHQ85_A)

NHIS <- NHIS %>%
  mutate(PHQ85_A = as.numeric(PHQ85_A),
         PHQ85_A = case_when(
           PHQ85_A== 1 ~ 1,
           PHQ85_A == 2 ~ 2,
           PHQ85_A== 3 ~ 3,
           PHQ85_A == 4 ~ 4,
           PHQ85_A == 7 ~ NA,
           PHQ85_A == 8 ~ NA,
           PHQ85_A == 9 ~ NA,
           TRUE ~ PHQ85_A))

table(NHIS$PHQ85_A)

 # PHQ86_A

 # feeling bad about yourself, or that you are a failure, or have let yourself or your family down

 # 1 - Not at all, 2 - Several Days, 3 - More than half the days, 4 - Nearly every day

table(NHIS$PHQ86_A)

NHIS <- NHIS %>%
  mutate(PHQ86_A = as.numeric(PHQ86_A),
         PHQ86_A = case_when(
           PHQ86_A== 1 ~ 1,
           PHQ86_A == 2 ~ 2,
           PHQ86_A== 3 ~ 3,
           PHQ86_A == 4 ~ 4,
           PHQ86_A == 7 ~ NA,
           PHQ86_A == 8 ~ NA,
           PHQ86_A == 9 ~ NA,
           TRUE ~ PHQ86_A))

table(NHIS$PHQ86_A)

 # PHQ87_A

 # trouble concentrating on things, such as reading the newspaper or watching television

 # 1 - Not at all, 2 - Several Days, 3 - More than half the days, 4 - Nearly every day

table(NHIS$PHQ87_A)

NHIS <- NHIS %>%
  mutate(PHQ87_A = as.numeric(PHQ87_A),
         PHQ87_A = case_when(
           PHQ87_A== 1 ~ 1,
           PHQ87_A == 2 ~ 2,
           PHQ87_A== 3 ~ 3,
           PHQ87_A == 4 ~ 4,
           PHQ87_A == 7 ~ NA,
           PHQ87_A == 8 ~ NA,
           PHQ87_A == 9 ~ NA,
           TRUE ~ PHQ87_A))

table(NHIS$PHQ87_A)


 # PHQ88_A

 # moving or speaking so slowly that other people could have noticed? Or the opposite, being so fidgety or restless that you have been moving around a lot more than usual?

 # 1 - Not at all, 2 - Several Days, 3 - More than half the days, 4 - Nearly every day

table(NHIS$PHQ88_A)

NHIS <- NHIS %>%
  mutate(PHQ88_A = as.numeric(PHQ88_A),
         PHQ88_A = case_when(
           PHQ88_A== 1 ~ 1,
           PHQ88_A == 2 ~ 2,
           PHQ88_A== 3 ~ 3,
           PHQ88_A == 4 ~ 4,
           PHQ88_A == 7 ~ NA,
           PHQ88_A == 8 ~ NA,
           PHQ88_A == 9 ~ NA,
           TRUE ~ PHQ88_A))

table(NHIS$PHQ88_A)


# PHQCAT_A

# How severe are your depression symptoms?

# 1 - None, 2 - Mild, 3 - Moderate, 4 - Severe

table(NHIS$PHQCAT_A)

NHIS <- NHIS %>%
  mutate(PHQCAT_A = as.numeric(PHQCAT_A),
         PHQCAT_A = case_when(
           PHQCAT_A== 1 ~ 1,
           PHQCAT_A == 2 ~ 2,
           PHQCAT_A== 3 ~ 3,
           PHQCAT_A == 4 ~ 4,
           PHQCAT_A == 7 ~ NA,
           PHQCAT_A == 8 ~ NA,
           PHQCAT_A == 9 ~ NA,
           TRUE ~ PHQCAT_A))

table(NHIS$PHQCAT_A)

# PHQSCORE

# Calculating a PHQ 'Score' for each individual to use as a measure for anxiety

NHIS$PHQSCORE <- rowMeans(NHIS[, c("PHQ81_A", "PHQ82_A", "PHQ83_A", "PHQ84_A", "PHQ85_A", "PHQ86_A", "PHQ87_A", "PHQ88_A")], na.rm = TRUE)

summary(NHIS$PHQSCORE)


#MDI - Mood Disorder Index

# combining the 'GADSCORE' metric and 'PHQSCORE' metric to build a an overall individual 'score' for mental health. Important to note that a higher score (max of 4) indicates poor mental health while a lower score (min of 1) indicates better mental health.

NHIS$MDI <- rowMeans(NHIS[, c("GADSCORE", "PHQSCORE")], na.rm = TRUE)

summary(NHIS$MDI)


```

## Non-Modifiable Independent Variables

### Sex

```{r}

#SEX_A, 

# ITEM : Sex of Sample Adult
# CODE: 1 - Male, 2 - Female, 7 - Refused, 8 - Not Ascertained, 9 - Don't Know

table(NHIS$SEX_A)

NHIS <- NHIS %>%
  mutate(SEX_A = as.numeric(SEX_A),
         SEX_A = case_when(
           SEX_A == 1 ~ ' Male',
           SEX_A == 2 ~ 'Female ',
           SEX_A == 7 ~ NA_character_,
           SEX_A == 8 ~ NA_character_,
           SEX_A == 9 ~ NA_character_,
           TRUE ~ as.character(SEX_A)),
         SEX_A = factor(SEX_A))

NHIS <- NHIS %>%
  rename(SEX = SEX_A)

table(NHIS$SEX)


```

### Age

```{r}

#AGEP_A

# ITEM : Age of selected adult

table(NHIS$AGEP_A)

# making the var more useful by translating it into four generational buckets

NHIS <- NHIS %>%
  mutate(AGEP_A = as.numeric(AGEP_A),
         AGEP_A = case_when(
           between(AGEP_A, 18, 25) ~ 'Generation Z',
           between(AGEP_A, 26, 39) ~ 'Millenial',
           between(AGEP_A, 40, 64) ~ 'Generation X',
           between(AGEP_A, 65, 85) ~ 'Baby Boomer',
           AGEP_A == 97 ~ NA_character_,
           AGEP_A == 98 ~ NA_character_,
           AGEP_A == 99 ~ NA_character_,
           TRUE ~ as.character(AGEP_A)),
         AGEP_A = factor(AGEP_A))

NHIS <- NHIS %>%
  rename(GEN = AGEP_A)

table(NHIS$GEN)

```

### Race

```{r}

# Race variable is HISPALLP_A
#HISPALLP_A

table(NHIS$HISPALLP_A)

NHIS <- NHIS %>%
  mutate(HISPALLP_A = as.numeric(HISPALLP_A),
         HISPALLP_A = case_when(
           HISPALLP_A == 1 ~ ' Hispanic ',
           HISPALLP_A == 2 ~ ' White ',
           HISPALLP_A == 3 ~ ' Black ',
           HISPALLP_A == 4 ~ ' Asian ',
           between(HISPALLP_A, 5, 7) ~ 'Other',
           HISPALLP_A == 7 ~ NA_character_,
           HISPALLP_A == 8 ~ NA_character_,
           HISPALLP_A == 9 ~ NA_character_,
           TRUE ~ as.character(HISPALLP_A)), 
         HISPALLP_A = factor(HISPALLP_A))

NHIS <- NHIS %>%
  rename( RACE= HISPALLP_A)

table(NHIS$RACE)


```

## Modifiable Independent Variables

### Exercise

```{r}

#MODTPR_A

# time period for moderate leisure time physical activities

# CODE - 0 - Never, 1 - Per day, 2 - Per week, 3 - Per month, 4 - Per year

table(NHIS$MODTPR_A)

NHIS <- NHIS %>%
  mutate(MODTPR_A = as.numeric(MODTPR_A),
         MODTPR_A = case_when(
           MODTPR_A == 0 ~ 1,
           MODTPR_A == 1 ~ 5,
           MODTPR_A == 2 ~ 4,
           MODTPR_A == 3 ~ 3,
           MODTPR_A == 4 ~ 2,
           MODTPR_A == 5 ~ NA,
           MODTPR_A == 6 ~ NA,
           MODTPR_A == 7 ~ NA,
           MODTPR_A == 8 ~ NA,
           MODTPR_A == 9 ~ NA,
           TRUE ~ MODTPR_A))

table(NHIS$MODTPR_A)

#VIGTPR_A

# frequency of vigorous physical activity times per week

# CODE - 0  Never, 1 - Per day, 2 - Per week, 3 - Per month, 4 - Per year

table(NHIS$VIGTPR_A)

NHIS <- NHIS %>%
  mutate(VIGTPR_A = as.numeric(VIGTPR_A),
         VIGTPR_A = case_when(
           VIGTPR_A == 0 ~ 1,
           VIGTPR_A == 1 ~ 5,
           VIGTPR_A == 2 ~ 4,
           VIGTPR_A == 3 ~ 3,
           VIGTPR_A == 4 ~ 2,
           VIGTPR_A == 5 ~ NA,
           VIGTPR_A == 6 ~ NA,
           VIGTPR_A == 7 ~ NA,
           VIGTPR_A == 8 ~ NA,
           VIGTPR_A == 9 ~ NA,
           TRUE ~ VIGTPR_A))

table(NHIS$VIGTPR_A)


# EXSCORE
# averaging individual respoded to VIGTPR and MODTPR to create an 'exercise score'

NHIS$EXSCORE <- rowMeans(NHIS[, c("MODTPR_A", "VIGTPR_A")], na.rm = TRUE)

summary(NHIS$EXSCORE)

```

### Sleep

```{r}
# SLEEP

#SLPHOURS_A

# On average, how many hours of sleep do you get in a 24-hour period?

# Range of values - 00 - 24 , 97-99 are NAs

table(NHIS$SLPHOURS_A)

NHIS <- NHIS %>%
  mutate(SLPHOURS_A = as.numeric(SLPHOURS_A),
         SLPHOURS_A = case_when(
           between(SLPHOURS_A, 0, 6) ~ 0,
           between(SLPHOURS_A, 7, 9) ~ 1,
           between(SLPHOURS_A,10, 24) ~ 0,
           SLPHOURS_A == 97 ~ NA,
           SLPHOURS_A == 98 ~ NA,
           SLPHOURS_A == 99 ~ NA,
           TRUE ~ SLPHOURS_A))

slphours_MDI <- aggregate(MDI ~ SLPHOURS_A, data = NHIS, FUN = mean)

print(slphours_MDI)

#SLPREST_A

# How often did you wake up feeling well-rested

# Code: 1 - Never, 2 - Some days, 3 - Most days, 4 - Every day

table(NHIS$SLPREST_A)

NHIS <- NHIS %>%
  mutate(SLPREST_A = as.numeric(SLPREST_A),
         SLPREST_A = case_when(
           SLPREST_A == 1 ~ 1,
           SLPREST_A == 2 ~ 2,
           SLPREST_A == 3 ~ 3,
           SLPREST_A == 4 ~ 4,
           SLPREST_A == 7 ~ NA,
           SLPREST_A == 8 ~ NA,
           SLPREST_A == 9 ~ NA,
           TRUE ~ SLPREST_A))

table(NHIS$SLPREST_A)

summary(NHIS$SLPREST_A)
```

### Diet

```{r}

# general approach is to re-code such that higher score is healthier

# DIET

# SODATP_A
 # Number of times drank soda - time period
 # CODE: 0 - Never, 1 - Day, 2 - Week, 3 - Month
NHIS <- NHIS %>%
  mutate(SODATP_A = as.numeric(SODATP_A),
         SODATP_A = case_when(
           SODATP_A == 0 ~ 4,
           SODATP_A == 1 ~ 1,
           SODATP_A == 2 ~ 2,
           SODATP_A == 3 ~ 3,
           SODATP_A == 7 ~ NA,
           SODATP_A == 8 ~ NA,
           SODATP_A == 9 ~ NA,
           TRUE ~ SODATP_A))

# FRIESTP_A
 # Number of times eat fried potatoes - time period
 # CODE: 0 - Never, 1 - Day, 2 - Week, 3 - Month

NHIS <- NHIS %>%
  mutate(FRIESTP_A = as.numeric(FRIESTP_A),
         FRIESTP_A = case_when(
           FRIESTP_A == 0 ~ 4,
           FRIESTP_A == 1 ~ 1,
           FRIESTP_A == 2 ~ 2,
           FRIESTP_A == 3 ~ 3,
           FRIESTP_A == 7 ~ NA,
           FRIESTP_A == 8 ~ NA,
           FRIESTP_A == 9 ~ NA,
           TRUE ~ FRIESTP_A))


# OVEGTP_A
 # Number of times eat other vegetables - time period
 # CODE: 0 - Never, 1 - Day, 2 - Week, 3 - Month

NHIS <- NHIS %>%
  mutate(OVEGTP_A = as.numeric(OVEGTP_A),
         OVEGTP_A = case_when(
           OVEGTP_A == 0 ~ 1,
           OVEGTP_A == 1 ~ 4,
           OVEGTP_A == 2 ~ 3,
           OVEGTP_A == 3 ~ 2,
           OVEGTP_A == 7 ~ NA,
           OVEGTP_A == 8 ~ NA,
           OVEGTP_A == 9 ~ NA,
           TRUE ~ OVEGTP_A))

#FRUITTP_A
# Number of times eat fruit - time period
# CODE: 0 - Never, 1 - Day, 2 - Week, 3 - Month

NHIS <- NHIS %>%
  mutate(FRUITTP_A = as.numeric(FRUITTP_A),
         FRUITTP_A = case_when(
           FRUITTP_A == 0 ~ 1,
           FRUITTP_A == 1 ~ 4,
           FRUITTP_A == 2 ~ 3,
           FRUITTP_A == 3 ~ 2,
           FRUITTP_A == 7 ~ NA,
           FRUITTP_A == 8 ~ NA,
           FRUITTP_A == 9 ~ NA,
           TRUE ~ FRUITTP_A))

# SALADTP_A
# Number of times eat salad 0 time period
# Code - 0 - Never, 1 - Day, 2 - Week, 3 - Month,

NHIS <- NHIS %>%
  mutate(SALADTP_A = as.numeric(SALADTP_A),
         SALADTP_A = case_when(
           SALADTP_A == 0 ~ 1,
           SALADTP_A == 1 ~ 4,
           SALADTP_A == 2 ~ 3,
           SALADTP_A == 3 ~ 2,
           SALADTP_A == 7 ~ NA,
           SALADTP_A == 8 ~ NA,
           SALADTP_A == 9 ~ NA,
           TRUE ~ SALADTP_A))

# Calculating a DIET 'Score' for each individual to use as a measure for diet

NHIS$DIET<- rowMeans(NHIS[, c("FRUITTP_A", "SALADTP_A", "FRIESTP_A", "OVEGTP_A", "SODATP_A")], na.rm = TRUE)

summary(NHIS$DIET)


```

### Alcohol

```{r}
# ALCOHOL

# DRK12MWK_A
# Days drank alcohol per week, past 12M
# CODE: 00 - Less than one day per week, 1-7 days per week, 95 - No drinks in last year

table(NHIS$DRK12MWK_A)

# keep 0 as 0, add '95' to 0
NHIS <- NHIS %>%
  mutate(DRK12MWK_A = as.numeric(DRK12MWK_A),
         DRK12MWK_A = case_when(
           DRK12MWK_A == 00 ~ 0,
           DRK12MWK_A == 95 ~ 0,
           DRK12MWK_A == 97 ~ NA,
           DRK12MWK_A == 98 ~ NA,
           DRK12MWK_A == 99 ~ NA,
           TRUE ~ DRK12MWK_A))

table(NHIS$DRK12MWK_A)

NHIS <- NHIS %>%
  rename(LIVER = DRK12MWK_A)

```

### Smoking

```{r}
# SMOKING

#SMKNOW_A
# Do you now smoke cigarettes every day, some days, or not at all
# CODE: 1- Every day. 2 - Some days, 3 - Not at all, 7 - 9 - NA
table(NHIS$SMKNOW_A)

NHIS <- NHIS %>%
  mutate(SMKNOW_A = as.numeric(SMKNOW_A),
         SMKNOW_A = case_when(
           SMKNOW_A == 1 ~ 3,
           SMKNOW_A == 2 ~ 2,
           SMKNOW_A == 3 ~ 1,
           SMKNOW_A == 7 ~ NA,
           SMKNOW_A == 8 ~ NA,
           SMKNOW_A == 9 ~ NA,
           TRUE ~ SMKNOW_A))

table(NHIS$SMKNOW_A)


# ECIGNOW_A
# Do you now use e-cigarettes or other electronic vaping products every day,
# some days, or not at all
# CODE: 1 - Every day, 2 - Some days, 3 - Not at all, 7,8,9 - NA
table(NHIS$ECIGNOW_A)

NHIS <- NHIS %>%
  mutate(ECIGNOW_A = as.numeric(ECIGNOW_A),
         ECIGNOW_A = case_when(
           ECIGNOW_A == 1 ~ 3,
           ECIGNOW_A == 2 ~ 2,
           ECIGNOW_A == 3 ~ 1,
           ECIGNOW_A == 7 ~ NA,
           ECIGNOW_A == 8 ~ NA,
           ECIGNOW_A == 9 ~ NA,
           TRUE ~ ECIGNOW_A))

table(NHIS$ECIGNOW_A)

# creating index of these two variables related to smoking 
NHIS$LUNGS<- rowMeans(NHIS[, c("ECIGNOW_A", "SMKNOW_A")], na.rm = TRUE)

summary(NHIS$LUNGS)

```

### Work

```{r}
# WORK

#EMPWKHRS3_A
# Hours worked per week
# Code: 01 - 95+ , 97 refused, 98 - Not Ascertained, 99 - Don't Know

table(NHIS$EMPWKHRS3_A)
summary(NHIS$EMPWKHRS3_A)

NHIS <- NHIS %>%
  mutate(EMPWKHRS3_A = as.numeric(EMPWKHRS3_A),
         EMPWKHRS3_A = case_when(
           between(EMPWKHRS3_A, 0, 39) ~  'Part Time',
           EMPWKHRS3_A == 40 ~ 'Full Time',
           between(EMPWKHRS3_A, 41, 95) ~ 'More than Full Time',
           EMPWKHRS3_A == 97 ~ NA,
           EMPWKHRS3_A == 98 ~ NA,
           EMPWKHRS3_A == 99 ~ NA,
           TRUE ~ as.character(EMPWKHRS3_A)), 
          EMPWKHRS3_A = factor(EMPWKHRS3_A))

table(NHIS$EMPWKHRS3_A)

NHIS <- NHIS %>%
  rename(WORK = EMPWKHRS3_A)


```

# Testing Relationships

## Corelation

```{r}
# checking correlations of dependent measures

 # mental health score and life satisfaction
cor.test(NHIS$MDI, NHIS$LSATIS4_A)

# of depression measures
cor.test(NHIS$PHQSCORE, NHIS$DEPFREQ_A, na.rm=T)

cor.test(NHIS$PHQSCORE, NHIS$MDI)

# of anxiety measures
cor.test(NHIS$GADSCORE, NHIS$ANXFREQ_A)

cor.test(NHIS$GADSCORE, NHIS$MDI)

# of anxieety and depression measures 
cor.test(NHIS$ANXFREQ_A, NHIS$DEPFREQ_A)

cor.test(NHIS$GADSCORE, NHIS$PHQSCORE)


ex_mh <- lm(MDI ~ EXSCORE, data = NHIS)

summary(ex_mh)

ex_mh2 <- lm(MDI ~ PA18_02R_A, data = NHIS)

summary(ex_mh2)

summary_activity_MDI <- NHIS %>%
  group_by(PA18_02R_A) %>%
  summarise(Average = mean(MDI, na.rm = TRUE),
            Variance = var(MDI, na.rm = TRUE))


summary_activity_MDI

# summarizing Sex variable
summary_sex_MDI <- NHIS %>%
  group_by(SEX) %>%
  summarise(Average = mean(MDI, na.rm = TRUE),
            Variance = var(MDI, na.rm = TRUE))

# Display the resulting table
print(summary_sex_MDI)

# summarizing GEN variable
summary_gen_MDI <- NHIS %>%
  group_by(GEN) %>%
  summarise(Average = mean(MDI, na.rm = TRUE),
            Variance = var(MDI, na.rm = TRUE))

# Display the resulting table
print(summary_gen_MDI)

# summarizing RACE variable
summary_race_MDI <- NHIS %>%
  group_by(RACE) %>%
  summarise(Average = mean(MDI, na.rm = TRUE),
            Variance = var(MDI, na.rm = TRUE))

# Display the resulting table
print(summary_race_MDI)


```

# Multivariable Modelling

## Non-Modifiable Risk Factors

```{r}
# Non-Modifiable Risk Factors

#overall mental health
nonmodif_fund_mod <- lm(MDI ~ SEX + GEN + RACE, data = NHIS)

summary(nonmodif_fund_mod)

#anxiety
anx_nonmodif1 <- lm(ANXFREQ_A ~ SEX + GEN + RACE, data = NHIS)

summary(anx_nonmodif1)

anx_nonmodif2 <- lm(GADSCORE ~ SEX + GEN + RACE, data = NHIS)
summary(anx_nonmodif2)


#depression
dep_nonmodif1 <- lm(DEPFREQ_A ~ SEX + GEN + RACE, data = NHIS)

summary(dep_nonmodif1)

dep_nonmodif2 <- lm(PHQSCORE ~ SEX + GEN + RACE, data = NHIS)
summary(dep_nonmodif2)


```

## Modifiable Risk Factors

```{r}
# Modifiable Fundamentals - Sleep, Diet, Exercise

# overall mental health
modif_fund_mod <- lm(MDI ~ SLPREST_A + EXSCORE + DIET,  data = NHIS)

summary(modif_fund_mod)

# Modifiable Extras

# overall mental health
modif_extras_mod <- lm(MDI ~ LIVER + LUNGS + WORK, data = NHIS)

summary(modif_extras_mod)
```

```{r}

# All Modfiables Risk Factors

# overall mental health
modif_all_mod <- lm(MDI ~  EXSCORE + DIET + SLPREST_A + LIVER + LUNGS + WORK, data = NHIS)

summary(modif_all_mod)

# depression

 # using DEPFREQ
dep_main_1 <- lm(DEPFREQ_A ~ SLPREST_A + EXSCORE + DIET + LIVER + LUNGS + WORK, data = NHIS)

summary(dep_main_1)

 # using PHQ
dep_main_2 <- lm(PHQSCORE ~ SLPREST_A + EXSCORE + DIET + LIVER + LUNGS +  WORK, data = NHIS)

summary(dep_main_2)

# anxiety

# using ANXFREQ
anx_main_1 <- lm(ANXFREQ_A ~ SLPREST_A + EXSCORE + DIET + LIVER + LUNGS + WORK, data = NHIS)

summary(anx_main_1)

 # using GAD
anx_main_2 <- lm(GADSCORE ~ SLPREST_A + EXSCORE + DIET + LIVER + LUNGS + WORK, data = NHIS)

summary(anx_main_2)
```

# Visualization

###Main Models

```{r}

# residuals plot of main models

plot(modif_all_mod)

plot(modif_fund_mod)

plot(nonmodif_fund_mod)


```

```{r}

# corrolelogram

# create smaller dataset with numeric variables
smaller <- select(NHIS,
                ANXFREQ_A,
                DEPFREQ_A,
                MDI,
                LSATIS4_A,
                GADSCORE,
                PHQSCORE,
                EXSCORE,
                DIET,
                SLPREST_A,
                LIVER,
                LUNGS,
               )

# set correlation
corr <- cor(smaller, use = "complete.obs")

# corrplot
corrplot(corr, 
        type = "lower", 
        method = "circle", 
        col = c("orchid", "white", "seagreen4"), 
        addCoef.col = "black", 
        tl.col = "black", 
        tl.srt = 80, 
        diag = FALSE,
        title = "Correlogram of NHIS",
        order = "hclust", 
        addrect = 8)


# total model coefficients
coefplot(modif_all_mod) + 
scale_x_continuous(limits = c(-1, 1))


# main model Rsquared values
r2_values<- data.frame(
  Model = c("Non-Modifiable Fundamentals", "Modifiable Fundamentals", "All Modifables Model"),
  R_squared = c(0.03, 0.18, 0.2) 
)


ggplot(r2_values, aes(x = Model, y = R_squared, fill = Model)) +
  geom_bar(stat = "identity", width = 0.5) +
  theme_minimal() +
  labs(title = "Comparison of R-squared Values", x = "Model", y = "R-squared Value") +
  scale_fill_brewer(palette = "Pastel1") 

```

```{r}

# Using Box-Plots to display differences in MH score across demographic variables

# Sex and Mental Health Score

theme_set(theme_classic())

MDIvSEX_plot <- ggplot(NHIS, aes(SEX, MDI))
MDIvSEX_plot + geom_boxplot(varwidth=T, fill="plum") + 
    labs(title="Box plot", 
         subtitle="MDI by Sex",
         caption="Source: MDI",
         x="Sex",
         y="MDI")

# Race

theme_set(theme_classic())

MDIvRACE_plot <- ggplot(NHIS, aes(RACE, MDI))
MDIvRACE_plot + geom_boxplot(varwidth=T, fill="plum") + 
    labs(title="Box plot", 
         subtitle="MDI by Race",
         caption="Source: MDI",
         x="Race",
         y="MDI")

# Age

theme_set(theme_classic())

AGEvRACE_plot <- ggplot(NHIS, aes(GEN, MDI))
AGEvRACE_plot + geom_boxplot(varwidth=T, fill="plum") + 
    labs(title="Box plot", 
         subtitle="MDI by Age",
         caption="Source: MDI",
         x="Generation",
         y="MDI")


```

# Interpretations / Results of Hypotheses

Base interpretations of our hypotheses, which will be expanded upon further particularly in the realm of practical significance in the later deliverable elements.

## Main

H1 : Modifiable factors will hold more predictive power for mental health outcomes than the non-modifiable factors

```{r}

# table comparing relevant models
stargazer(nonmodif_fund_mod, modif_all_mod,
 type = "text",
 add.lines = list(c("Model Name", "Non-Modifiable", "Modifiable")))


```

Results: H1 is confirmed by the fact that the R-squared values of the modifiables (.2) is higher than that of non-modifiables (.03).

H2 : Modifiable basics such as sleep, exercise, and diet will have the more predictive power than the additional modifiable independent variables.

```{r}
# table comparing relevant models
stargazer( modif_fund_mod, modif_extras_mod,
 type = "text",
 add.lines = list(c("Model Name", "Modifiable Fundamentals", "Modifiable Extras")))

```

Results: H2 is confirmed as the R-squared value for the modif_fund_mod (.18) is higher than that of the modif_extras_mod (.008).

## Modifiable Factors

H3: Higher exercise scores are associated with lower score on MDI - as EXSCORE increases, GADSCORE and PHQSCORE will decrease, the effect size will be greater for depression than anxiety.

```{r}
cor.test(NHIS$EXSCORE, NHIS$GADSCORE)

cor.test(NHIS$EXSCORE, NHIS$PHQSCORE)

# table comparing relevant models
stargazer( modif_all_mod, dep_main_2, anx_main_2,
 type = "text",
 add.lines = list(c("Model Name", "All Modifiables", "Depression Modifiables", "Anxiety Modifiables")))


```

Results: H3 is not confirmed, EXSCORE beta coefficient is only significant in GADSCORE model and in the opposite direction expected for each the MDI model, PHQSCORE model and GADSCORE model.

H4: Higher sleep scores associated with better mental health outcomes, more for depression than anxiety.

```{r}

cor.test(NHIS$SLPREST_A, NHIS$GADSCORE)

cor.test(NHIS$SLPREST_A, NHIS$PHQSCORE)

# table comparing relevant models
stargazer( modif_all_mod, dep_main_2, anx_main_2,
 type = "text",
 add.lines = list(c("Model Name", "All Modifiables", "Depression Modifiables", "Anxiety Modifiables")))


```

Results: H4 confirmed, Sleep variable is significant in each MDI, PHQ, and GAD model each in the expected direction.

H5: Higher diet scores associated with lower score on MDI, GAD and PHQ, effect greater for depression than anxiety.

```{r}

cor.test(NHIS$DIET, NHIS$GADSCORE)

cor.test(NHIS$DIET, NHIS$PHQSCORE)

# table comparing relevant models
stargazer( modif_all_mod, dep_main_2, anx_main_2,
 type = "text",
 add.lines = list(c("Model Name", "All Modifiables", "Depression Modifiables", "Anxiety Modifiables")))

```
Results:H5 confirmed, 'DIET' index significant in MDI, PHQSCORE, and GADSCORE models, each in the expected direction.

H6: For depression, a stronger association with drinking than smoking.

```{r}

cor.test(NHIS$LUNGS, NHIS$PHQSCORE)

cor.test(NHIS$LIVER, NHIS$PHQSCORE)

# table comparing relevant models
stargazer( dep_main_1, dep_main_2,
 type = "text",
 add.lines = list(c("Model Name", "Depression Frequency Model", "PHQ Depression Model")))


```

Results:H6 not confirmed, 'LIVER' index is not significant in either depression model, whereas 'LUNGS' index is significant and in expected direction. 

H7: For anxiety, a stronger association with smoking than drinking.

```{r}

cor.test(NHIS$LUNGS, NHIS$GADSCORE)

cor.test(NHIS$LIVER, NHIS$GADSCORE)

# table comparing relevant models
stargazer( anx_main_1, anx_main_2,
 type = "text",
 add.lines = list(c("Model Name", "Anxiety Frequency Model", "GAD Model")))


```

Results: H7 confirmed. In both anxiety models, 'LUNGS' variable is significant and in the expected direction, whereas 'LIVER' variable is not significant.

H8: Weekly hours of work beyond full time are associated with lower mental health scores.

```{r}

# table to compare relevant models
stargazer( modif_all_mod, dep_main_2, anx_main_2,
 type = "text",
 add.lines = list(c("Model Name", "All Modifiables", "Depression Modifiables", "Anxiety Modifiables")))

```

Results: H8 not confirmed, beta coefficient for 'Work More than Full Time' is not significant in any of the MDI, PHQ, or GAD models. Coefficient for 'Work Part Time' is significant for each model and associated with a slightly higher score on each metric. 

## Non-Modifiable Factors

H9: Gen Z with least favorable (highest) score on MDI, GAD, PHQ,. More favorable with each increase in generation.

```{r}

stargazer( nonmodif_fund_mod, anx_nonmodif2, dep_nonmodif2,
 type = "text",
 add.lines = list(c("Model Name", "Non-Modifiable MDI", "Non-Modifiable Dep", "Non-Modifiable Anx")))

```

Results: H9 confirmed. There seems to an inverse relationships in our models between Age or generation and mental health score. As the generation and age decrease, the beta coefficients in our MDI, GAD, and PHQ models increase, indicating poorer mental health. These effects are statistically significant and greater for anxiety than depression.

H10: Female group will have worse average mental health score, these differences more pronounced for depression than anxiety.

```{r}
# table comparing relevant models
stargazer(nonmodif_fund_mod, anx_nonmodif2, dep_nonmodif2,
 type = "text",
 add.lines = list(c("Model Name", "Non-Modifiable Fundamentals", "Anxiety Nonmodifiables", "Depression Nonmodifiables")))

```

Results: H10 partially confirmed, beta coefficient for Female dummy variable is in expected direction and significant in MDI, PHQ, and GAD models. However, the effect size is slightly smaller for GADSCORE than PHQSCORE.

H11: No significant differences in average mental health scores across racial groups.

```{r}

# table comparing relevant models
stargazer(nonmodif_fund_mod, anx_nonmodif2, dep_nonmodif2,
 type = "text",
 add.lines = list(c("Model Name", "Non-Modifiable Fundamentals", "Anxiety Nonmodifiables", "Depression Nonmodifiables")))


```

Results: H11 not confirmed. There are statistically significant differences between racial groups in all three non-modifiable models - that for MDI, GAD, and PHQ. 

```{r}

# table with confirmed and not confirmed hypotheses

# Creating a data frame with Hypothesis # and Confirmation Status
hypotheses_data <- data.frame(
  `Result` = c("Confirmed", "Confirmed", "Not Confirmed", "Confirmed", "Confirmed", "Not Confirmed", "Confirmed", "Not Confirmed", "Not Confirmed", "Partially Confirmed", "Not Confirmed" ) 
)

# Printing the created data frame
print(hypotheses_data)


```

# Limitations

Some ideas for limitations which will be discussed fully in the final deliverables (Presentation and Policy Memo)

*diet variable* Still considering the changing of or removing of the main diet variable (DIET), as this is a rather limited index for capturing the complexity of a quality diet model. Unfortunately, there are significant limitations from the survey design in this area which make it difficult to form a quality index variable for diet.

*causal uncertainty* The old 'correlation does not mean causation' quip is relevant here. This isn't to say that correlation is meaningless here (especially because of the potentially cyclical nature of these relationships), but rather to state up front that interpretations of many of our models are limited in terms of causal inference because of a various considerations - they are non-longitudinal, potential for confounding variables or omitted variable bias, and reverse causality.
